---
title: "EDS241: Assignment 2"
author: "Erica Bishop"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, 
                      warning = FALSE, tidy.opts = list(width.cutoff = 60), 
                      tidy = TRUE)

#devtools::install_github('rstudio/rmarkdown')
options(scipen=999) # not scientific notation

```

```{r load_packages}

#load libraries 
library(tidyverse)
library(estimatr)
library(stargazer)
library(janitor)
library(here)
library(corrplot)

```

```{r read_data}

smoking_data <- read_csv(here("eds241_data/SMOKING_EDS241.csv")) |> 
  clean_names()

```

# Question 1: Application of estimators based on the "treatment of ignorability" assumption

The goal is to estimate the causal effect of maternal smoking during pregnancy on infant birth
weight using the treatment ignorability assumptions (Lecture 6 & 7). The data are taken from the
National Natality Detail Files, and the extract “SMOKING_EDS241.csv”' is a random sample of
all births in Pennsylvania during 1989-1991. Each observation is a mother-infant pair. The key
variables are:

**The outcome and treatment variables are:**
birthwgt=birth weight of infant in grams
tobacco=indicator for maternal smoking

**The control variables are:**
mage (mother's age), meduc (mother's education), mblack (=1 if mother black), alcohol (=1 if
consumed alcohol during pregnancy), first (=1 if first child), diabete (=1 if mother diabetic),
anemia (=1 if mother anemic)


## (a) 
What is the unadjusted mean difference in birth weight of infants with smoking and 
nonsmoking mothers? Under what assumption does this correspond to the average treatment effect 
of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical 
evidence for or against this assumption.

```{r}
#calculate difference in means

means <- smoking_data |> 
  group_by(tobacco == 1) |> 
  summarise(mean(birthwgt, na.rm = TRUE))

mean_smoking <- as.numeric(means[2,2])

mean_nonsmoking <- as.numeric(means[1,2])

mean_diff_unadj <- mean_nonsmoking - mean_smoking

```

```{r results = 'asis', echo = F}
#print statements
print(paste("The unadjusted difference in average birthweights between babies born to mothers that smoked and mothers that did not smoke is", round(mean_diff_unadj, 2), "grams."))

```

This unadjusted mean difference only corresponds to the average treatment effect under the treatment ignorability assumption, which states that there are no other variables that influence both the treatment (smoking and non-smoking) and the outcome (birth weight). Some evidence against this assumption is demonstrated in the correlation matrix below, as there are multiple variables that are correlated to both the treatment and the outcome, for example age has a 0.14 correlation with tobacco usage and a 0.1 correlation with birth weight. Education is also correlated with both birth weigth and tobacco usage. 

```{r}

#show correlations between other variables
cormatrix <- cor(smoking_data)

corrplot(cormatrix,
         method = "shade",
         shade.col = NA,
         tl.col = "black",
         tl.srt = 45,
         addCoef.col = "black",
         number.font = 8,
         cl.pos = "n",
         order = "FPC")


```



## (b) 
Assume that maternal smoking is randomly assigned conditional on the observable
covariates listed above. Estimate the effect of maternal smoking on birth weight using an OLS
regression with linear controls for the covariates. Report the estimated coefficient on tobacco
and its standard error.

```{r}
#run OLS
ols_mod <- lm(data = smoking_data,
              formula = birthwgt ~ tobacco
              + mage
              + meduc
              + mblack
              + alcohol
              + first 
              + diabete
              + anemia)

ols_results <- summary(ols_mod)

ols_results

```

```{r results = 'asis', echo = F}

print(paste0("The coefficient on tobacco is ", round(ols_results$coefficients[2,1], 3), ". This means that if all other variables are held constant, the difference in infants born to smoking mothers would be about 228 grams lighter than babies born to mothers that didn't smoke."))

print(paste0("The standard error for the effect of tobacco in this model is ", round(ols_results$coefficients[2,2], 3),"."))

```


## (c) 
Use the exact matching estimator to estimate the effect of maternal smoking on birth weight.
For simplicity, consider the following covariates in your matching estimator: create a 0-1
indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if
meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). These 4
covariates will create 2*2*2*2 = 16 cells. Report the estimated average treatment effect of
smoking on birthweight using the exact matching estimator and its linear regression analogue.

```{r}

#encode data into bins by specified thresholds
smoking_encoded <- smoking_data |> 
  mutate(mage = as.factor(
    case_when(mage >= 34 ~ 1, #greater than 34 = 1
                          mage < 34 ~ 0)), #less than 34 = 0
         meduc = as.factor(
           case_when(meduc >= 16 ~ 1, #greater than 16 = 1
                           meduc < 16 ~ 0)),# less than 16 = 0
    mblack = as.factor(mblack),
    alcohol = as.factor(alcohol)
    ) %>% 
  select(c(mage, meduc, mblack, alcohol, tobacco, birthwgt)) %>%  #drop other variables
  add_column(X = as.factor(case_when( #create a factor for 16 unique combinations of covariate
                 .$meduc == 0 & .$mage == 0 & .$mblack == 0 & .$alcohol == 0 ~ 1,
                 .$meduc == 1 & .$mage == 0 & .$mblack == 0 & .$alcohol == 0 ~ 2,
                 .$meduc == 1 & .$mage == 1 & .$mblack == 0 & .$alcohol == 0 ~ 3,
                 .$meduc == 1 & .$mage == 0 & .$mblack == 1 & .$alcohol == 0 ~ 4,
                 .$meduc == 1 & .$mage == 0 & .$mblack == 0 & .$alcohol == 1 ~ 5,
                 .$meduc == 0 & .$mage == 1 & .$mblack == 0 & .$alcohol == 0 ~ 6,
                 .$meduc == 0 & .$mage == 1 & .$mblack == 1 & .$alcohol == 0 ~ 7,
                 .$meduc == 0 & .$mage == 1 & .$mblack == 0 & .$alcohol == 1 ~ 8,
                 .$meduc == 0 & .$mage == 0 & .$mblack == 1 & .$alcohol == 0 ~ 9,
                 .$meduc == 0 & .$mage == 0 & .$mblack == 1 & .$alcohol == 1 ~ 10,
                 .$meduc == 0 & .$mage == 0 & .$mblack == 0 & .$alcohol == 1 ~ 11,
                 .$meduc == 1 & .$mage == 1 & .$mblack == 1 & .$alcohol == 1 ~ 12,
                 .$meduc == 0 & .$mage == 1 & .$mblack == 1 & .$alcohol == 1 ~ 13,
                 .$meduc == 1 & .$mage == 0 & .$mblack == 1 & .$alcohol == 1 ~ 14,
                 .$meduc == 1 & .$mage == 1 & .$mblack == 0 & .$alcohol == 1 ~ 15,
                 .$meduc == 1 & .$mage == 1 & .$mblack == 1 & .$alcohol == 0 ~ 16
                 )))


TIA_table <- smoking_encoded %>%
  group_by(X,tobacco)%>% 
  summarise(n_obs = n(),
            Y_mean = mean(birthwgt, na.rm = TRUE))%>% #Calculate number of observations and Y mean by X by treatment cells
  ungroup()%>%
  mutate(total_obs = sum(n_obs))%>% #Calculate total number of observations
  group_by(tobacco)%>%
  mutate(total_obs_d = sum(n_obs))%>% #Calculate total number of observations by treatment cells
  group_by(X)%>%
  mutate(Y_diff = lead(Y_mean)-Y_mean,
         W_ATE = sum(n_obs)/total_obs,
         W_ATT = lead(n_obs)/lead(total_obs_d))%>% #Calculate difference in outcome and ATE and ATT weights by X
  ungroup()%>%
  mutate(ATE=sum(W_ATE*Y_diff, na.rm= TRUE),
         ATT=sum(W_ATT*Y_diff, na.rm= TRUE))%>% #Calculate ATE and ATT 
  mutate_if(is.numeric, round, 2) #Round data

# TIA_table

gt::gt(TIA_table)


```

```{r}
# MULTIVARIATE MATCHING AS REGRESSION ESTIMATOR
se_models = starprep(ols_mod, stat = c("std.error"), se_type = "HC2", alpha = 0.05)
stargazer(ols_mod, se = se_models, type="text")
```


```{r results = 'asis', echo = F}

print(paste("The estimated average treatment effect of smoking on birthweight is", round(TIA_table[1, 10], 3), "grams, using the exact matching method."))

print(paste("This differs only slightly (by about 3 grams) from the estimated effect of smoking on infant birthweigth using the linear regression analogue, which was", round(ols_results$coefficients[2,1], 3), "grams."))

```





